---
title: "Overwatch Hero Pick Rates"
author: "Alex Montgomery"
date: "31/04/2021"
output: html_document
---

```{r libraries, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) #load tidyverse
library(plotly)    #load plotly for interactive graphing
```

```{r overwatch logo, out.width = "50%", fig.align = "center", echo=FALSE}
#Load and Display the Overwatch logo
img1 <- "images/overwatch_logo.png"
knitr::include_graphics(img1)
```

## Overwatch
Overwatch is a team-based multiplayer first-person shooter with a player base of around **10 million players** per month. Each match sees a uniquely chosen set of heroes go head to head in a 6v6 battle. Players can choose from **32 heroes** belonging to 3 main classes: tank, damage and support. Each team is allowed two heroes from each class. 

The composition of heroes on a team greatly alters from game to game based on many variables such as personal preference, opposition hero picks, synergy with other heroes and the **skill level of players**. Different heroes require different skillsets, with some skillsets taking longer to develop than others. 

Overwatch is played competitively with a sophisticated skill rating system with **7 divisions** ranging from Bronze (lowest) to Grandmaster (highest). **[Skill rating](https://dotesports.com/overwatch/news/overwatch-sr-calculation-explained-16535)** (SR) has a large range from 0 to 5000 and determines an individuals division placement. The formula for determining SR is mostly kept secret but generally players lose or gain SR based on whether they win or lose a game. Due to the sophistication of the SR system an individuals SR, and therefore division, are a very accurate representation of their ability in the game. 

## Current Study
As previously mentioned, how often specific heroes are picked partly depends on the skill level of players. The current study shall set out to use visualisations to explore the relationship between **hero pick rate** and **skill level (division)**. The visualisations should be able to identify which specific heroes are picked more in lower ranks than higher ranks and vice versa, as well as be able to identify heroes whos pick rate does not significantly vary based on skill level. 

## Data Origin
The data set used in this study was manually scraped from [Overbuff.com](https://www.overbuff.com/heroes). It includes data from the 3 months before it was scraped (March 24th) on competitive Overwatch played on PC. The data set includes all **32 heroes** and their average pick rate, win rate, tie rate and on fire rates in each of the 7 divisions. Average **pick rates** in each of the divisions are the only columns of interest for the present study, meaning the data will need cleaning. In this dataset, pick rate refers to the amount a hero is picked as a percentage of all the heroes. This means that when adding up the percentage pick rate of all the 32 heroes in one division the total will be 100%.

```{r overbuff visualisation logo, out.width = "70%", fig.align = "center", echo=FALSE}
#Load and Display the Overwatch logo
img2 <- "images/overbuff_visualisation.png"
knitr::include_graphics(img2)
```

The way the data is presented by [Overbuff](https://www.overbuff.com/heroes) makes it difficult to compare **hero pick rates** across divisions. Each division has to be clicked and separately viewed. This would not be difficult to compare across divisions if there was only around 5 heroes. However, their are **32 heroes**, making it very difficult to see trends of heroes picked more often in different divisions. 

```{r load data, echo=TRUE, warning=FALSE, include=TRUE}
#Load the data set
df = read.csv("hero_stats_3_months.csv")
#Check the data is as expected
head(df, 3)
```

## Visualisation 1 - Stacked Bar Chart Comparing Pick Rate Across All 7 Divisions
Visualisation 1 shall set out to use all the data available for a more accurate image of how hero pick rate varies across the 7 divisions. To do this a stacked bar chart will be used. 

A stacked bar chart will show a stand out trend of what heroes are picked more overall across all divisions. A more detailed look at the visualisation will show the differences in pick rates of each hero across the 7 divisions. Using plotly to create the visualisation (instead of ggplot) allows the use of the hover label feature. This feature is especially useful with stacked bar charts as many bars do not have a 0 starting point making it hard to determine the value that the bar represents.

## Cleaning data for Visualisation 1
As previously mentioned the only columns of interest to this study are the hero pick rates in each division. The data was first cleaned by selecting only these columns and the hero column. A new column was created to facilitate the ordering of hero by total pick rate in visualisation 1. The dataset for visualisation 1 was then ordered by total pick rate. Normally this step would not be needed as plotly has a reorder function that can change the order of the x axis. However, adding traces in a loop overrides this function meaning this step was necessary to be performed in the data cleaning process.

```{r clean data for p1, echo=TRUE, warning=FALSE, include=TRUE}
#Transform original data set to only include the heroes and pick rates for each division
dfp1 <- df %>% select(1, 2, 6, 10, 14, 18, 22, 26)
#Create a variable that is the sum of pick rate across all divisions so that visualisaion 1 can be ordered as desired
dfp1$totalpick <- rowSums(dfp1[,c(2:8)])
#Transform dfp1 to arrange heroes by total pick rate to create the desired order in the visualisation
dfp1 <- dfp1 %>% arrange(desc(totalpick))
#Check the data is as expected
head(dfp1, 5)
```

## Creating the Visualisation

```{r create visualisation 1, echo=TRUE, warning=FALSE, include=TRUE}
#Set the plot space and add the first division
p1 <- plot_ly(dfp1, x = ~hero,
                    y = ~bronze_pick_rate, 
                    name = "Bronze Pick Rate",
                    type = "bar", 
                    width = 900,
                    height = 600,
                    #Set the colour of the first division and create a black line around each bar
                    marker = list(color = "#E0B5AC",
                                  line = list(color = "#000000, 0.5)",
                                  width = 1))) %>%

  #Set the layout specifications of the visualisation
  layout(title = "Hero Pick Rate by Division",
                      #Set the order of the heroes to be determined by the order they appear in the dataset
         xaxis = list(categoryarray = ~hero, 
                      categoryorder = "array",
                      #Set the x axis title and the angle and size of the labels for each hero
                      title = "Hero",
                      tickangle =270),
                      tickfont = list(size = 10),
                      #Set the y axis title, scale format, range and type of bar chart
         yaxis = list(title = "Pick Rate", 
                      tickformat = "%",
                      range = c(0, 0.70)),
                      barmode = "stack")
#Create a value containing the 6 divisions not yet plotted for use in a for loop
  divplot <- colnames(dfp1[c(3:8)])
#Create a value for use in the add trace loop to set the colours for each division (colours match the exact hex codes that Overbuff assigned to each division)
  colours <- c("#CAD6CC", "#EFEB8B", "#4E98BB", "#D9E8F3", "#EDA152", "#FEE072")
#Create a loop to add a trace for each of the remaining divisions
  for(i in divplot){
    p1 <- p1 %>% add_trace(x = dfp1[["hero"]], 
                           y = dfp1[[i]],
                           #Replace "_" with " " and capitalise the division titles
                           name = gsub(x = i, pattern = "\\_", replacement = " ") %>% str_to_title(i),
                           #Change the colours of the divisions per the "colours" variable previously created
                           marker = list(color = colours[[match(i, divplot)]]))
  }
#Display p1 visualisation  
p1
```

## Visualisation 1 Summary
Visualisation 1 allows comparison of the pick rates across all divisions, however, there are some limitations of the design. Firstly, the cumulative pick rate of a hero does demonstrate which heroes are picked more overall. However, it does not represent a real world figure as the overall average pick rate for a hero should be the average across all 7 divisions rather than the sum (as it is represented in the visualisation). A grouped bar chart would have combated this issue. However, this would include too many bars on the x axis as each hero would have 7 bars, totaling 224 bars. 

A further issue of visualisation 1 is that it is hard to see stand out trends due to the overwhelming amount of data that is being shown. In hindsight the inclusion of all the data may have been more detrimental than useful in regards to the readability of the visualisation. Upon closer inspection it is still possible to identify heroes with significantly varying pick rates across divisions, such as Moira being picked far more in lower divisions compared to higher ones. However, it would take some time to examine the whole visualisation to pick out all the significant trends. 

This issue shall be addressed in the creation of visualisation 2. Less data shall be represented in the visualisation but with the important trends still being identifiable and much more easily spotted, with less time needed to examine the visualisation.

## Visualisation 2 - Bump Chart Comparing Rank Order of Heroes in the Bronze and Grandmsater Divsions
Upon inspection of visualisation 1 it seems that the relationship between skill level (division) and pick rate is mostly linear, meaning that no heroes were barely picked in the highest and lowest divisions but picked a lot in the middle divisions. Because of this it was thought appropriate to omit the data of the middle divisions and just compare the data from the highest and lowest divisions. This simplifies the visualisation but to further simplify the visualisation heroes were ranked by their pick rate in each division and directly compared. A bump chart perfectly facilitates this comparison and so was deemed the best type of chart to use. 

## Cleaning data for visualiation 2

As mentioned above, visualisation 2 shall only focus on the highest and lowest divisions, so the hero, bronze pick rate and grandmaster pick rate columns were first selected. It was necessary to create a new dataset as the data for this visualisation needed to be in a long format as opposed to a wide format in order to facilitate the ranking of heroes. Rankings were then created for each division by grouping heroes by division and then ordering heroes by pick rate and assigning a rank based on their row number.

```{r clean data for p2, echo=TRUE, warning=FALSE, include=TRUE}
#Transform the original data set to only include the heroes and pick rates for the lowest and highest divisions
dfp2 <- df %>% select(1, 2, 26)
#Pivot the data from wide to long to facilitate the ranking of heroes
dfp2 <- gather(dfp2, division, pick_rate, bronze_pick_rate, grandmaster_pick_rate, factor_key = TRUE)
#Create a new variable that ranks heroes based on their pick rate in each division
dfp2 <- dfp2 %>%
        group_by(division) %>%
        arrange(division, desc(pick_rate), hero) %>% 
        mutate(rank = row_number()) %>%
        ungroup()
#Check the data is as expected
head(dfp2, 5)
```

## Creating the Visualisation

```{r create visualisation 2, fig.width=8.5 ,fig.height=9, echo=TRUE, warning=FALSE, include=TRUE}


#Set the plot place
p2 <- ggplot(dfp2, aes(x = division, y = rank, group = hero)) +
      #Add a line for each hero and make it slighlty transparent for better visability of crossing lines
      geom_line(aes(color = hero, alpha = 1), size = 2) +
      #Add a point for each hero to better clarify the start and end point of each line
      geom_point(aes(color = hero, alpha = 1), size = 4) +
      #Add a white down in the middle of each point
      geom_point(color = "#FFFFFF", size = 1) +
      #Create geom_text to facilitate adding heroes names by their bronze pick rate ranking
      geom_text(data = dfp2 %>% filter(division == "bronze_pick_rate"),
            #Add heroes names by their bronze pick rate ranking    
            aes(label = hero, x = 0.8) , hjust = 0.5,
            #Make the font bold, black and a speciifc size
            fontface = "bold", color = "#000000", size = 4) +
      #Create geom_text to facilitate adding heroes names by their grandmaster pick rate ranking
      geom_text(data = dfp2 %>% filter(division == "grandmaster_pick_rate"),
            #Add heroes names by their grandmaster pick rate ranking    
            aes(label = hero, x = 2.2) , hjust = 0.5,
            #Make the font bold, black and a speciifc size
            fontface = "bold", color = "#000000", size = 4) +
      #Specify tick labels on the x axis
      scale_x_discrete(labels = c("Bronze Pick Rate Rank", "Grandmaster Pick Rate Rank")) +
      #Reverse the y axis scale to have the top heroes as the most picked and specify the scale to go up in increments of 1 up to 32
      scale_y_reverse(breaks = 1:32) + 
      #Specify the size of the plot title and make it bold
      theme(plot.title = element_text(size=16, face = "bold"),
            #Specify the size of the tick labels on the x axis
            axis.text.x = element_text(size=12),
            #Specify the size of the axis titles and make them bold
            axis.title = element_text(size=14,face = "bold"),
            #Change the paneling on the plot to make it easy to compare across the visualisation
            panel.grid.major.y = element_blank(),
            #Remove the legend 
            legend.position = "none") +
      #Specify the labels of the x and y axis
      labs(x = "Division", 
           y = "Hero Rank", 
           #Specify the title and subtitle
           title = "Overwatch Hero Pick Rate",
           subtitle = "Heroes ranked by % pick rate in the lowest and highest divisions (Bronze and Grandmaster) over the last 3 months") +
      #Specify the colours of the heroes lines to match the colour associated with each hero in Overwatch 
      scale_color_manual(values = c("#2178E9", "#67121D", "#3EA2CB", "#6E994D", "#F3B23A", "#FF7FD1", "#DDCCAC", "#04ADEF", "#84FE01", "#938848", "#D39308", "#8BEC22", "#C83C3C", "#9ADBF4", "#FFE16C", "#B3365D", "#BCC84A", "#2178E9", "#A49D86", "#AA958E", "#C19477", "#5DAB81", "#5D81EB", "#A07DC8", "#5CECFF", "#FF6200", "#F8911B", "#7878D4", "#8991A6", "#4D7883", "#F571A8", "#C79C00")) 
#Display p2 visualisation
p2
```

## Summary of Visualisation 2
Overall, visualisation 2 makes it far easier to spot trends of heroes that are picked significantly more or less in higher or lower divisions. The steeper the slope of the line, the bigger the difference in pick rate there is from the lowest division to the highest division. It is relatively easy to pick out the steepest slopes to identify the heroes whos pick rate is the most effected by skill level (division). The opposite is true when identifying heroes whos pick rate is least affected by skill level, just look for the flattest lines. Picking out specific heroes is made easier by the choice of colour for the lines. Each heroes line is represented by the colour that is associated with that specific hero. With the target demographic reading this visualisation being individuals who are interested in Overwatch, it is safe to assume that they will know what colours are associated with each hero. To further make the visualisation more readable the lines were made slightly transparent to combat the issue of many lines crossing one another.

## Project Summary
Although visualisation 1 was deemed too busy to easily pick out trends it was a necessary step to take to arrive at the final visualisation (#2). Visualisation 1 demonstrated that too much data made the visualisation more difficult to interpret. Visualisation 1 also showed that the relationship between skill level (division) and pick rate was mostly linear meaning that the data for the middle divisions was not necessary to include in the final visualisation.

The final visualisation is more than adequate for the purpose it was created, to identify trends in **hero pick rate** across **skill level**. Multiple heroes can be easily identified as being picked significantly more at higher skill levels. These heroes are identified by the steepest lines going up to the right and include heroes such as Zenyatta, Wrecking Ball, Tracer, Bridgette and Widowmaker. Multiple heroes can be easily identified as being picked significantly more at lower skill levels. These heroes can be identified by the steepest lines going down to the right and include heroes such as Moira, Junkrat, Soldier:76, Orisa, Genji and Pharah. It is also possible to identify heroes which are generally preffered by both low and high skill players as they are at the top of the rankings for both divisions. These heroes include Mercy, Reinhardt, Zarya and Mcree.

Overall the final visualisation presents an adequate amount of data that makes the visualisation very easy to read whilst still providing valuable insights into the relationship between skill level and hero pick rate for each of the 32 heroes in Overwatch.




